Omics-based Single Step Trait Prediction

### 2017-09-15 paper/manuscript.tex
updates after aem confus

% -   Objectives (Hybrids)
In animal breeding, single step prediction, particularly H-BLUP, has already 
become a routine procedure~\cite{Legarra2014} but, to the best of our knowledge, 
has only been considered in plants in a study on wheat~\cite{Ashraf2016} and
white spruce lines, respectively~\cite{Ratcliffe2017}.
While single step prediction has been applied to pig hybrids~\cite{Xiang2016}
this application is not equivalent to hybrid breeding in plants.
There, the maternal and the paternal crossing partner can be selected without
constraints whereas animal breeders select only males because of the limited
fertility of the female animals.
Technological advances in plant hybrid breeding such as the rapid generation
of completely homozygous lines~\cite{Wedzony2009} now enable breeders to
produce about $n = 1,000$ individuals per generation and sex.
Assuming that about 90\% of these lines were never before tested in any hybrid
combination about $n^{2} * 0.9^{2} = 810,000$ putative hybrids per generation
would not have a single parent with phenotypically tested progeny
\cite{Westhues2017}.
Predicting these hybrid candidates with low relatedness to tested hybrids using
only information from their parents is currently a hot topic in plant breeding
\cite{Technow2014,Kadam2016,Westhues2017}.
Hence, our first objective was to examine whether the single step prediction 
framework can be transfered to a situation where a very large number of hybrids
($\approx$ 1 million) needs to be predicted.

% -   Objectives (quantitative predictores)
Given that the availability of pedigree and genomic data is the most prevalent
situation in animal breeding, previous studies have focused exclusively on the
application of single step prediction to these predictors.
Recent investigations into the utility of `omics' data for boosting predictive
abilities in plants have suggested that metabolites and transcriptomic data can
improve upon predictive abilities based on pedigree or genomic data
\cite{Guo2016,Xu2016,Dan2016,Zenke-Philippi2017,Westhues2017}.
Yet, these technologies are not fully mature and therefore incur higher costs
to breeders than pedigree and genomic data, putting their utility into question
despite their proven usefulness.
Our second objective is therefore to evaluate the impact of covering only a
subset of all genotypes with both, the complete and the incomplete predictor on
predictive ability, where for the remaining individuals only genomic data are
available.

\section{Materials and Methods}
\Genetics2level{Exp1}
To address the second objective a maize inbred line data set, subsequently
referred to as Exp1, was used~\cite{Yang2014}.
Exp1 originally comprised 513 maize lines representing the global maize
diversity  and was reduced to the set of tropical and subtropical lines
($n = 211$) given that it is the largest of the four pre-classified
subgroups~\cite{Yang2014}.
All inbred lines were evaluated in five Chinese environments described in detail
by~\citeNP{Yang2014}.
Best linear unbiased predictors (BLUPs) were calculated for 17
traits~\cite{Guo2016} of which six are presented in this study
(Table~\ref{table:TraitKey}).
The selection of these traits was done so that four traits were predicted with
greater precision by transcriptomic data and two further traits were predicted
with greater precision by genomic data.
MaizeSNP50 BeadChip-data were available for the entire set of 513 lines from the
diversity panel.
Additionally, RNA sequencing was performed for 368 lines on young seedlings 15 
days after sowing.
By exploiting identity by descent (IBD) between 49,728 SNPs on the BeadChip 
overlapping with the RNA-seq data, 556,809 high quality SNPs could be inferred 
for these lines~\cite{Fu2013,Li2013}.
For the remaining 145 maize lines without RNA-seq data, high density markers
were inferred via projection of IBD regions onto the BeadChip, using this core 
set of markers~\cite{Yang2014}.
After filtering for minor allele frequency ($\geq 5$\%), heterozygosity rate
($\leq 5$\%) and call frequency ($\geq 95$\%), missing genotypes were imputed 
for 37,760 SNPs using the Beagle software~\cite{Browning2009}.
Gene expression data were normalized using a normal quantile transformation to
satisfy modeling assumptions~\cite{Fu2013}.
A subset of 28,850 annotated genes was available for 149 out of the 211
genotypes~\cite{Fu2013} and provided the basis for further analyses.

@@

\Genetics2level{Population structure}
Population structure in Exp1 was examined through a STRUCTURE
\cite{Pritchard2000} analysis with the number of putative ancestral
populations $K$ ranging from 2 to 9.
Based on the estimated probability of the data for each $K$, we chose $K=4$
and carried out a principal component analysis (PCA) on the scaled and centered
genomic feature matrix using the \emph{pca} function from the \emph{R}-package
\emph{LEA}~\cite{Frichot2015}.
Least-squares estimates of ancestry proportions were estimated based on sparse 
nonnegative matrix factorization algorithms~\cite{Frichot2014} using the 
\emph{snmf} function in \emph{LEA}.
Genotypes in the PCA plot (Fig.~\ref{fig:inbred-pca}) were colorized based on their
primary assignment to any of the four putative ancestral populations.
To mitigate the possibility of inflated predictive abilities, which can arise
from large differences in the average performance of the different
subpopulations~\cite{Windhausen2012}, we reduced the panel of inbred lines to
those assigned to either cluster 1 or cluster 4 in Fig.~\ref{fig:inbred-pca}.
This reduction yielded 164 genotypes, 110 of which were covered by both, genomic
and transcriptomic data.

@@

\Genetics2level{Prediction}
\subsection{Kernels}
BLUP models are computationally attractive when the number of features $p$
exceeds the number of genotypes $n$ and are equivalent to a selection index
when fixed effects have been accomodated by the dependent variable 
\cite{Mrode2014}.
Depending on the data set, up to three predictors were available for agronomic
trait predictions, namely pedigree data (P), genomic data (G) and
transcriptomic data (T).
The corresponding feature matrix of the inbred lines ($\mathbf{M}$) has
dimensions $n \times p$ where $n$ pertains to the number of genotypes.
For the sake of clarity, we do not distinguish between models for Exp1 and Exp2
at this point.
It should be noted, however, that in the case of Exp2 the term `genotype' refers
strictly to the parent lines.
For the complete predictor we generated kernels by centering and standardizing
all features in $\mathbf{M}$ to unit variance:

@@

where $M$ denotes the number of features~\cite{VanRaden2008}.
In the case of pedigree data (P), coancestry coefficients were used directly
for $\mathbf{K}$.

\subsection{Genetic model}
The general model for genetic values was:

@@

\subsection{Single step prediction}
If the genetic values are fully captured by the features in $\mathbf{M}$,
Eq.~\ref{eq:KBLUPModel} can be expressed as

\begin{align} \label{eq:KBLUPModelSimplified}
	\mathbf{y} &= \mu + \mathbf{Z} \mathbf{g} + \boldsymbol{e} \\
	&= \mu + \mathbf{Z} \mathbf{M} \boldsymbol{\alpha} + \boldsymbol{e},
\end{align}

@@

Let us further denote the matrix of centered covariates of the incomplete
predictor as
\begin{equation} \label{eq:incomplete-covariates}
\tensor[_{I}]{\mathbf{W}}{} =
\begin{bmatrix}
  \mathbf{Z}_{1} \; \tensor[_{I}]{\mathbf{\hat{W}}}{_{1}} \\
  \tensor[_I]{\mathbf{W}}{_{2}}
 \end{bmatrix},
\end{equation}
where observed feature covariates are stored in $\tensor[_I]{\mathbf{W}}{_{2}}$ while feature
covariates for genotypes not covered by the incomplete predictor
($\tensor[_{I}]{\mathbf{\hat{W}}}{_{1}} = \tensor[_C]{\mathbf{K}}{_{12}} \; \tensor[_C]{\mathbf{K}}{_{22}^{-1}} \; \tensor[_I]{\mathbf{W}}{_{2}}$)
need to be imputed.
In this case we need to partition the vector of genetic values $\mathbf{g}$
into a component $\mathbf{g}_{1}$ (\textit{i.e.,} genetic values of genotypes
covered only by the complete predictor) and a component
$\mathbf{g}_{2} = \tensor[_{I}]{\mathbf{W}}{_{2}} \; 
\tensor[_{I}]{\boldsymbol{\alpha}}{}$ (\textit{i.e.,} genetic
values of genotypes covered by both predictors where $\tensor[_{I}]{\boldsymbol{\alpha}}{}$
pertains to the random feature effects of the incomplete predictor).
Under multivariate normality, the vector $\mathbf{g}_1$ can be written as the
sum of the conditional expectation given $\mathbf{g}_2$ and a residual error
term:
\begin{align} \label{eq:mrna1}
	\mathbf{g_1} &= E(\mathbf{g}_1|\mathbf{g}_2) + \boldsymbol{\epsilon} \\
	&= \tensor[_{C}]{\mathbf{K}}{_{12}} \; \tensor[_{C}]{\mathbf{K}}{_{22}^{-1}} \; \tensor[_{I}]{\mathbf{W}}{_{2}} \;
	\tensor[_{I}]{\boldsymbol{\hat{\alpha}}} + (\mathbf{g_1} - \tensor[_{C}]{\mathbf{K}}{_{12}} \;
	\tensor[_{C}]{\mathbf{K}}{_{2}^{-1}} \;
	\tensor[_{I}]{\boldsymbol{\hat{\alpha}}}{}) \\
	&= \mathbf{\hat{g}}_1 + \boldsymbol{\epsilon},
\end{align}

where $\tensor[_{C}]{\mathbf{K}}{_{ij}}$ are partitions of
$\tensor[_{C}]{\mathbf{K}}{}$ that correspond to $\mathbf{g}_{1}$ and
$\mathbf{g}_{2}$, respectively~\cite{Fernando2014} and where
$\mathbf{\hat{g}}_{1}$ pertains to the imputed genetic values of genotypes
not equipped with the incomplete predictor.

@@

and the covariance of $\mathbf{\hat{g}}_{1}$ is
$\tensor[_{C}]{\mathbf{K}}{_{12}} \;
\tensor[_{C}]{\mathbf{K}}{_{22}^{-1}} \;
\tensor[_{I}]{\mathbf{W}}{_{2}} \;
\tensor[_{I}]{\mathbf{W}}{_{2}^{\top}} \;
\tensor[_{C}]{\mathbf{K}}{_{22}^{-1}} \;
\tensor[_{C}]{\mathbf{K}}{_{21}}\sigma^{2}_{\alpha}$
\cite{Fernando2014}.

The structure of the residual imputation/prediction error is known and can 
therefore be modeled.

@@

\begin{equation} \label{eq:single-step-model}
\mathbf{y} =
\mathbf{Xb} + \tensor[_{I}]{\mathbf{W}}{} \; \tensor[_{I}]{\boldsymbol{\alpha}} + \mathbf{U} \boldsymbol{\epsilon} + \mathbf{e},
\end{equation}

@@

Following the notation in \citeNP{Fernando2014} the phenotypes can be untangled 
as:

\begin{align} \label{eq:entangled-augmented-single-step-model}
\begin{bmatrix}
  \mathbf{y}_1 \\
  \mathbf{y}_2
 \end{bmatrix}
& =
 \begin{bmatrix}
  \mathbf{X}_1 \\
  \mathbf{X}_2
 \end{bmatrix}
 \boldsymbol{\beta} +
 \begin{bmatrix}
  \mathbf{Z}_1 & \boldsymbol{0} \\
  \boldsymbol{0} & \mathbf{Z}_2
 \end{bmatrix}
\begin{bmatrix}
  \mathbf{g}_1 \\
  \mathbf{g}_2
 \end{bmatrix}
  + \mathbf{e} \\
    & =
 \begin{bmatrix}
 \mathbf{X}_1 \\
 \mathbf{X}_2
 \end{bmatrix}
 \boldsymbol{\beta} +
 \begin{bmatrix}
  \mathbf{Z}_1 & \boldsymbol{0} \\
  \boldsymbol{0} & \mathbf{Z}_2
 \end{bmatrix}
\begin{bmatrix}
  \tensor[_{C}]{\mathbf{K}}{_{12}} \; \tensor[_{C}]{\mathbf{K}}{_{22}^{-1}} \;
  \tensor[_{I}]{\mathbf{W}}{_{2}} \;
  \tensor[_{I}]{\boldsymbol{\alpha}} + \boldsymbol{\epsilon}  \\
  \tensor[_{I}]{\mathbf{W}}{_{2}} \; \tensor[_{I}]{\boldsymbol{\alpha}} \\
 \end{bmatrix}
  + \mathbf{e} \\
    & =
 \begin{bmatrix}
  \mathbf{X}_1 \\
  \mathbf{X}_2
 \end{bmatrix}
 \boldsymbol{\beta} +
 \begin{bmatrix}
  \mathbf{Z}_1 & \boldsymbol{0} \\
  \boldsymbol{0} & \mathbf{Z}_2
 \end{bmatrix}
\begin{bmatrix}
  \tensor[_{I}]{\mathbf{\hat{W}}}{_{1}} \; \tensor[_{I}]{\boldsymbol{\alpha}} + \boldsymbol{\epsilon} \\
  \tensor[_{I}]{\mathbf{W}}{_{2}} \; \tensor[_{I}]{\boldsymbol{\alpha}} \\
 \end{bmatrix}
  + \mathbf{e} \\
\end{align}

The above equation implies that ultimately single step prediction uses only
one random effect vector $\mathbf{g}$, which entails the observed effects of
the incomplete predictor $\mathbf{g}_{2}$ and the imputed effects
$\mathbf{\hat{g}}_{1}$ for genotypes that are not equipped with the incomplete
predictor plus the residual imputation error $\boldsymbol{\epsilon}$.

@@

\begin{align} \label{eq:final-model}
\mathbf{y} &= \mathbf{Xb} +
\mathbf{Z} \; \tensor[_{I}]{\mathbf{W}}{} \;
\tensor[_{I}]{\boldsymbol{\alpha}} +
\mathbf{Z}\mathbf{U} \boldsymbol{\epsilon} +
\mathbf{e}.
\end{align}

Extending this model to hybrids with parent lines from two heterotic groups
simply requires separating the feature matrices of the two predictors into a
paternal and a maternal component, respectively.
Using the index $D$ for the Dent and the index $F$ for the Flint lines in Exp2
we then have a feature matrix $\mathbf{M}_{D}$ and $\mathbf{M}_{F}$ for each
predictor.
For the hybrids the final model is therefore simply
\begin{align} \label{eq:final-model}
\mathbf{y} &= \mathbf{Xb} +
\mathbf{Z}_{D} \;
\tensor[_{I}]{\mathbf{W}}{_{D}} \; \tensor[_{I}]{\boldsymbol{\alpha}}{_{D}} +
\mathbf{Z}_{D} \mathbf{U}_{D} \boldsymbol{\epsilon}_{D} +
\mathbf{Z}_{F} \;
\tensor[_{I}]{\mathbf{W}}{_{F}} \;
\tensor[_{I}]{\boldsymbol{\alpha}}{_{F}} +
\mathbf{Z}_{F}\mathbf{U}_{F} \boldsymbol{\epsilon}_{F} +
\mathbf{e}.
\end{align}

Breeding values or, separately for each heterotic group, general combining
abilities can be estimated as:

@@

For the hybrid data, the obtained GCA effects for the inbred lines represent 
half their breeding values.
The predicted agronomic performance of the inbred lines is

@@

and that of the hybrids is equivalently

\begin{equation} \label{eq:inbred-predicted-performance}
\hat{\mathbf{y}} = \mathbf{Z}_{D}\mathbf{\hat{g}}_{D} + \mathbf{Z}_{F}\mathbf{\hat{g}}_{F}.
\end{equation}

@@

For the hybrid data, LOOCV was carried out as follows:
Let $D$ and $F$ denote the set of parental inbred lines from the Dent and the 
Flint group, respectively.
Further, let $H \cap [D \times F]$ denote the set of hybrids from crosses 
between $D$ and $F$.
Note that $H$ does not comprise all possible crosses between $D$ and $F$,
which would have been $142 \times 103 = 14,626$ hybrids, but a subset of 1,521
hybrids.
Let $H_{i \times j}$ denote the hybrid to be predicted.
We can then define the parents of the training set hybrids as
$D_{i} = D \setminus \{i\}$ and $F_{j} = F \setminus \{j\}$, respectively.
The training set is then defined as 
$H_{TRN}(i \times j) = H \cap (D_{i} \times F_{j})$.

We judged the performance of each model by looking at its predictive ability,
which is calculated as $\rho(\mathbf{y}, \mathbf{\hat{y}})$, where 
$\mathbf{\hat{y}}$ is the vector of predicted values for all genotypes.

@@

\Genetics2level{Core sampling}
In order to evaluate the influence of the fraction of genotypes that is not
covered by the incomplete predictor on predictive ability we generated core
sets of nine different sizes from Exp1 (Fig.~\ref{fig:compound-plot}~a).
The base of each of these core sets was the superset of genotypes that were
covered by both, genomic and trancriptomic data.
A core set was created by artificially declaring a pre-specified fraction 
$x = {10\%, 20\%, \dots, 90\%}$ of genotypes as lacking of transcriptomic 
information.
For each value of $x$ we created 100 core sets $s \in {1, 2, 3, \dots, 100}$ by
repeatedly sampling genotypes, whose trancsriptomic information was declared
as missing, at random (Fig.~\ref{fig:compound-plot}~b).
The artificially removed transriptomic information on these genotypes was then
imputed.

@@

\section*{Results}
\Genetics2level{Exp1}
At the level of the 110 genotypes for which both, genomic (G) and 
transcriptomic (T) information, were available, the use of T yielded higher
predictive abilities than the use of G for two out of six traits
(Fig.~\ref{fig:InbredResults}A).
In terms of predictive ability, the absolute advantage of T over G was as large 
as 0.15 as observed for the trait ED\@.
The predictive ability for predictions based on G were always higher for the
full set of inbred lines ($n = 164$) than for the reduced set of inbred lines
($n = 110$).
The combination of G and T (\textit{i.e.}, GT) in a single-step approach for the 
prediction of all 164 lines, performed at least as well as G alone when T was 
superior over G for the reduced set of 110 lines.
In the case of KW and ED, the use of GT yielded a slight improvement in predictive 
ability over G.
Where T was a worse predictor for the reduced set of 110 lines than G,
predictive abilities of GT for the full set of 164 lines were also worse than 
those based on G alone.

\begin{figure}[H]
  \centering
  \includegraphics{./tables_figures/pred_ability_inbred.pdf}
  \caption{
  Predictive abilities and average, bootstrapped standard errors for
  tropical/subtropical inbred lines from the maize diversity panel and six
  agronomic traits.
  As predictors, genomic (G) data, transcriptomic (T) data and their 
  combination (GT) were used.
  The `Reduced' panel includes a subset of 110 lines covered by genomic and
  transcriptomic information whereas the `Full' panel comprises all 164
  genotypes.
  In the `Full' set of genotypes, transcriptomic records were imputed for 54
  lines.
  }
\label{fig:InbredResults}
\end{figure}

>>>>

% -   Hybrids
In animal breeding, single step prediction, particularly H-BLUP, has already 
become a routine procedure~\cite{Legarra2014} but, to the best of our knowledge, 
has only been considered in plants in a study on wheat~\cite{Ashraf2016} and
white spruce lines, respectively~\cite{Ratcliffe2017}.
Moreover, while an application of the single step framework to animals exists
for pigs~\cite{Xiang2016}, it is yet lacking for plant hybrids.
Technological advances in plant hybrid breeding such as the rapid generation
of completely homozygous lines~\cite{Wedzony2009} now enable breeders to
produce about $n = 1,000$ individuals per generation and sex.
Assuming that about 90\% of these lines were never before tested in any hybrid
combination about $n^{2} * 0.9^{2} = 810,000$ putative hybrids per generation
would not have a single parent with phenotypically tested progeny
\cite{Westhues2017}.
Predicting these hybrid candidates with low relatedness to tested hybrids using
only information from their parents is currently a hot topic in plant breeding
\cite{Technow2014,Kadam2016,Westhues2017}.

% -   quantitative predictores
Given that the availability of pedigree and genomic data is the most prevalent
situation in animal breeding, previous studies have focused exclusively on the
application of single step prediction to these predictors.
Recent investigations into the utility of `omics' data for boosting predictive
abilities in plants have suggested that metabolites and transcriptomic data can
improve upon predictive abilities based on pedigree or genomic data
\cite{Guo2016,Xu2016,Dan2016,Zenke-Philippi2017,Westhues2017}.
Yet, these technologies are not fully mature and therefore incur higher costs
to breeders than pedigree and genomic data, putting their utility into question
despite their proven usefulness.

% - Objectives
Hence, our objectives were to i) examine whether the single step prediction 
framework can be transfered to a situation where a very large number of hybrids
($\approx$ 1 million) needs to be predicted and ii) evaluate the impact of
covering only a subset of all genotypes with both, the complete and the
incomplete predictor on predictive ability, where for the remaining individuals
only genomic data are available.

\section{Materials and Methods}
\Genetics2level{Exp1}
To address the second objective a maize inbred line data set, subsequently
referred to as Exp1, was used~\cite{Yang2014}.
Exp1 originally comprised 513 maize lines representing the global maize
diversity  and was reduced to the set of tropical and subtropical lines
($n = 211$) given that it is the largest of the four subgroups classified by
\citeN{Yang2014}.
All inbred lines were evaluated in five Chinese environments described in detail
by~\citeNP{Yang2014}.
Best linear unbiased predictors (BLUPs) were calculated by~\citeN{Yang2014} for
17 traits~\cite{Guo2016} of which six are presented in this study
(Table~\ref{table:TraitKey}).
The selection of these traits was done so that four traits were predicted with
greater precision by transcriptomic data and two further traits were predicted
with greater precision by genomic data.
MaizeSNP50 BeadChip-data were available for the entire set of 513 lines from the
diversity panel.
Additionally, RNA sequencing was performed for 368 lines on young seedlings 15 
days after sowing.
By exploiting identity by descent (IBD) between 49,728 SNPs on the BeadChip 
overlapping with the RNA-seq data, 556,809 high quality SNPs could be inferred 
for these lines~\cite{Fu2013,Li2013}.
For the remaining 145 maize lines without RNA-seq data, high density markers
were inferred via projection of IBD regions onto the BeadChip, using this core 
set of markers~\cite{Yang2014}.
After filtering for minor allele frequency ($\geq 5$\%), heterozygosity rate
($\leq 5$\%) and call frequency ($\geq 95$\%), missing genotypes were imputed 
for 37,760 SNPs using the Beagle software~\cite{Browning2009}.
Gene expression data were normalized using a normal quantile transformation to
satisfy modeling assumptions~\cite{Fu2013}.
A subset of 28,850 annotated genes was available for 149 out of the 211
genotypes~\cite{Fu2013} and provided the basis for further analyses.

@@

\Genetics2level{Population structure}
Population structure in Exp1 was examined through a STRUCTURE
\cite{Pritchard2000} analysis with the number of putative ancestral
populations $K$ ranging from 2 to 9.
Based on the estimated probability of the data for each $K$, we chose $K=4$
and carried out a principal component analysis (PCA) on the scaled and centered
genomic feature matrix using the \emph{pca} function from the \emph{R}-package
\emph{LEA}~\cite{Frichot2015}.
Least-squares estimates of ancestry proportions were estimated based on sparse 
nonnegative matrix factorization algorithms~\cite{Frichot2014} using the 
\emph{snmf} function in \emph{LEA}.
Genotypes in the PCA plot (Fig.~\ref{fig:inbred-pca}) were colorized based on their
primary assignment to any of the four putative ancestral populations.
To mitigate the possibility of inflated predictive abilities, which can arise
from large differences in the average performance of the different
subpopulations~\cite{Windhausen2012}, we reduced the panel of inbred lines to
those assigned to the largely overlapping clusters 1 and 4 in
Fig.~\ref{fig:inbred-pca}.
This reduction yielded 164 genotypes, 110 of which were covered by both, genomic
and transcriptomic data.

@@

\Genetics2level{Prediction}
\subsection{Kernels}
Depending on the data set, up to three predictors were available for agronomic
trait predictions, namely pedigree data (P), genomic data (G) and
transcriptomic data (T).
The corresponding feature matrix of the inbred lines ($\mathbf{M}$) has
dimensions $n \times p$ where $n$ pertains to the number of genotypes and $p$
pertains to the number of features; that is transcripts for T and SNPs for G.
For the sake of simplicity the derivations are detailed first for predicting
breeding values of inbred lines.
Later, we briefly describe the extension to models required for predicting the
genetic value of hybrids.
For the complete predictor we generated kernels by centering and standardizing
all features in $\mathbf{M}$ to unit variance:

@@

where $M$ denotes the number of features~\cite{VanRaden2008}.
In the case of pedigree data (P), the numerator relationship matrix was used
directly for $\mathbf{K}$.

\subsection{Genetic model}
The general model for breeding values of the lines was:

@@

\subsection{Single step prediction}
If the breeding values are fully captured by the features in $\mathbf{M}$,
Eq.~\ref{eq:KBLUPModel} can be expressed as

\begin{align} \label{eq:KBLUPModelSimplified}
	&= \mu + \mathbf{Z} \mathbf{M} \boldsymbol{\alpha} + \boldsymbol{e},
\end{align}

@@

Let us further denote the matrix of centered covariates of the incomplete
predictor as
\begin{equation} \label{eq:incomplete-covariates}
\tensor[_{I}]{\mathbf{W}}{} =
\begin{bmatrix}
  \mathbf{Z}_{1} \; \tensor[_{I}]{\mathbf{\hat{W}}}{_{1}} \\
  \tensor[_I]{\mathbf{W}}{_{2}}
 \end{bmatrix},
\end{equation}
where observed feature covariates are stored in $\tensor[_I]{\mathbf{W}}{_{2}}$
while the feature covariates for genotypes not covered by the incomplete
predictor can be estimated/imputed with the aid of the covariance matrix of the
complete predictor $\tensor[_{C}]{\mathbf{K}}{}$ and the feature matrix of the
incomplete predictor $\tensor[_{I}]{\mathbf{W}}{_{22}}$ by calculating
($\tensor[_{I}]{\mathbf{\hat{W}}}{_{1}} = \tensor[_C]{\mathbf{K}}{_{12}} \;
\tensor[_C]{\mathbf{K}}{_{22}^{-1}} \; \tensor[_I]{\mathbf{W}}{_{2}}$).
In this case we need to partition the vector of genetic values $\mathbf{g}$
into a component $\mathbf{g}_{1}$ (\textit{i.e.,} genetic values of genotypes
covered only by the complete predictor) and a component
$\mathbf{g}_{2} = \tensor[_{I}]{\mathbf{W}}{_{2}} \; 
\tensor[_{I}]{\boldsymbol{\alpha}}{}$ (\textit{i.e.,} genetic
values of genotypes covered by both predictors where $\tensor[_{I}]{\boldsymbol{\alpha}}{}$
pertains to the random feature effects of the incomplete predictor).
Under multivariate normality, the vector $\mathbf{g}_1$ can be written as the
sum of the conditional expectation given $\mathbf{g}_2$ and a residual error
term:
\begin{align} \label{eq:mrna1}
	\mathbf{g_1} &= E(\mathbf{g}_1|\mathbf{g}_2) +
	\boldsymbol{\epsilon} \\
	&= \tensor[_{C}]{\mathbf{K}}{_{12}} \;
       	\tensor[_{C}]{\mathbf{K}}{_{22}^{-1}} \;
       	\tensor[_{I}]{\mathbf{W}}{_{2}} \;
	\tensor[_{I}]{\boldsymbol{\hat{\alpha}}}{} +
       	(\mathbf{g_1} - \tensor[_{C}]{\mathbf{K}}{_{12}} \;
	\tensor[_{C}]{\mathbf{K}}{_{22}^{-1}} \;
       	\tensor[_{I}]{\mathbf{W}}{_{2}} \;
	\tensor[_{I}]{\boldsymbol{\hat{\alpha}}}{}) \\
	&= \mathbf{\hat{g}}_1 + \boldsymbol{\epsilon},
\end{align}

where $\tensor[_{C}]{\mathbf{K}}{_{ij}}$ are partitions of the matrix
$\tensor[_{C}]{\mathbf{K}}{}$ that correspond to $\mathbf{g}_{1}$ and
$\mathbf{g}_{2}$, respectively~\cite{Fernando2014} and where
$\mathbf{\hat{g}}_{1}$ pertains to the imputed genetic values of genotypes
not equipped with the incomplete predictor.

@@

and the covariance of $\mathbf{\hat{g}}_{1}$ is
$\tensor[_{C}]{\mathbf{K}}{_{12}} \;
\tensor[_{C}]{\mathbf{K}}{_{22}^{-1}} \;
\tensor[_{I}]{\mathbf{W}}{_{2}} \;
\tensor[_{I}]{\mathbf{W}}{_{2}^{\top}} \;
\tensor[_{C}]{\mathbf{K}}{_{22}^{-1}} \;
\tensor[_{C}]{\mathbf{K}}{_{21}} \;
\tensor[_{I}]{\sigma^{2}_{\alpha}}{}$
\cite{Fernando2014}.

The structure of the residual imputation/prediction error is known and,
therefore, can modeled.

@@

\begin{equation} \label{eq:single-step-model}
\mathbf{y} =
\mathbf{Xb} +
\tensor[_{I}]{\mathbf{W}}{} \;
\tensor[_{I}]{\boldsymbol{\alpha}}{} +
\mathbf{U} \boldsymbol{\epsilon} +
\mathbf{e},
\end{equation}

@@

According to the notation in \citeNP{Fernando2014} the phenotypes can be
untangled as:

\begin{align} \label{eq:entangled-augmented-single-step-model}
\begin{bmatrix}
  \mathbf{y}_1 \\
  \mathbf{y}_2
 \end{bmatrix}
& =
 \begin{bmatrix}
  \mathbf{X}_1 \\
  \mathbf{X}_2
 \end{bmatrix}
 \boldsymbol{\beta} +
 \begin{bmatrix}
  \mathbf{Z}_1 & \boldsymbol{0} \\
  \boldsymbol{0} & \mathbf{Z}_2
 \end{bmatrix}
\begin{bmatrix}
  \mathbf{g}_1 \\
  \mathbf{g}_2
 \end{bmatrix}
  + \mathbf{e} \\
    & =
 \begin{bmatrix}
 \mathbf{X}_1 \\
 \mathbf{X}_2
 \end{bmatrix}
 \boldsymbol{\beta} +
 \begin{bmatrix}
  \mathbf{Z}_1 & \boldsymbol{0} \\
  \boldsymbol{0} & \mathbf{Z}_2
 \end{bmatrix}
\begin{bmatrix}
  \tensor[_{C}]{\mathbf{K}}{_{12}} \; \tensor[_{C}]{\mathbf{K}}{_{22}^{-1}} \;
  \tensor[_{I}]{\mathbf{W}}{_{2}} \;
  \tensor[_{I}]{\boldsymbol{\alpha}}{} + \boldsymbol{\epsilon}  \\
  \tensor[_{I}]{\mathbf{W}}{_{2}} \; \tensor[_{I}]{\boldsymbol{\alpha}} \\
 \end{bmatrix}
  + \mathbf{e} \\
    & =
 \begin{bmatrix}
  \mathbf{X}_1 \\
  \mathbf{X}_2
 \end{bmatrix}
 \boldsymbol{\beta} +
 \begin{bmatrix}
  \mathbf{Z}_1 & \boldsymbol{0} \\
  \boldsymbol{0} & \mathbf{Z}_2
 \end{bmatrix}
\begin{bmatrix}
  \tensor[_{I}]{\mathbf{\hat{W}}}{_{1}} \;
  \tensor[_{I}]{\boldsymbol{\alpha}}{} + \boldsymbol{\epsilon} \\
  \tensor[_{I}]{\mathbf{W}}{_{2}} \; \tensor[_{I}]{\boldsymbol{\alpha}} \\
 \end{bmatrix}
  + \mathbf{e} \\
\end{align}

The model above indicates that single step prediction models ultimately use
only the matrix of known and imputed feature covariates $\mathbf{W}$ for the
estimation of breeding values while accounting for the imputation error
$\boldsymbol{\epsilon}$.

@@

\begin{align} \label{eq:final-model}
\mathbf{y} &= \mathbf{Xb} +
\mathbf{Z} \;
\tensor[_{I}]{\mathbf{W}}{} \;
\tensor[_{I}]{\boldsymbol{\alpha}}{} +
\mathbf{Z}\mathbf{U} \boldsymbol{\epsilon} +
\mathbf{e}.
\end{align}

Breeding values can be estimated as:

@@

The predicted agronomic performance of the inbred lines is

@@

The previous models can simply be extended to hybrids by defining separate
feature matrices for each predictor and each heterotic group; here
$\mathbf{M}_{D}$ for Dent and $\mathbf{M}_{F}$ for Flint parent lines in Exp2.
The general model for the calculation of genetic values in Exp2 is then

\begin{equation} \label{eq:hybrid-genetic-value-model}
  \mathbf{y} = \mu + \mathbf{Z}_{D} \mathbf{g}_{D} +
  \mathbf{Z}_{F} \mathbf{g}_{F} +
  \mathbf{e},
\end{equation}

corresponding to Eq.~\ref{eq:KBLUPModel}.

@@

For the hybrid data, LOOCV was carried out as follows:
Let $D$ and $F$ denote the set of parental inbred lines from the Dent and the 
Flint group, respectively.
The set $H$ is the subset of 1,521 phenotyped hybrids within the set of all
possible hybrids $D \times F$, which would comprise $14,626 = 142 * 103$
hybrids.
Let $i \times j$ denote the hybrid to be predicted.
We can then define the parents of the training set hybrids as
$D_{i} = D \setminus \{i\}$ and $F_{j} = F \setminus \{j\}$, respectively.
The training set is then defined as 
$H_{TRN}(i \times j) = H \cap (D_{i} \times F_{j})$.

We judged the performance of each model by looking at its predictive ability,
which was calculated as the correlation $\rho(\mathbf{y}, \mathbf{\hat{y}})$,
where $\mathbf{\hat{y}}$ is the vector of predicted values for all genotypes.

@@

\Genetics2level{Core sampling}
In order to evaluate the influence of the fraction of genotypes that is not
covered by the incomplete predictor on predictive ability we generated core
sets of nine different sizes from Exp1 (Fig.~\ref{fig:compound-plot}~a).
The base of each of these core sets was the set of 110 genotypes that were
covered by both, genomic and trancriptomic data.
A core set was created by artificially declaring a pre-specified fraction 
$x = {10\%, 20\%, \dots, 90\%}$ of genotypes as lacking of transcriptomic 
information.
For each value of $x$ we created 100 core sets $s \in {1, 2, 3, \dots, 100}$ by
repeatedly sampling genotypes, whose trancsriptomic information was declared
as missing, at random (Fig.~\ref{fig:compound-plot}~b).
The artificially removed transriptomic information on these genotypes was then
imputed.
For the each core set, comprising 110 genotypes, predictive abilities were again
computed as described above, using LOOCV and bootstrapping.

@@

\section*{Results}
\Genetics2level{Exp1}
At the level of the 110 genotypes for which both, genomic (G) and 
transcriptomic (T) information were available, the use of T yielded higher
predictive abilities than the use of G for two out of six traits
(Fig.~\ref{fig:InbredResults}A).
In terms of predictive ability, the absolute advantage of T over G was as large 
as 0.15 as observed for the trait ED\@.
The predictive ability for predictions based on G were always higher for the
full set of inbred lines ($n = 164$) than for the reduced set of inbred lines
($n = 110$).
The combination of G and T (\textit{i.e.}, $\tensor[_{C}]{G}{}
\tensor[_{I}]{T}{}$, where the subscripts correspond to the complete ($C$) and
the incomplete ($I$) predictor, respectively) in a single-step approach for the 
prediction of all 164 lines, performed at least as well as G alone when T was 
superior over G for the reduced set of 110 lines.
In the case of KW and ED, the use of $\tensor[_{C}]{G}{}\tensor[_{I}]{T}{}$
yielded a slight improvement in predictive ability over G.
Where T was a worse predictor for the reduced set of 110 lines than G,
predictive abilities of $\tensor[_{C}]{G}{}\tensor[_{I}]{T}{}$ for the full set
of 164 lines were also worse than those based on G alone.
